---
title: "Random Forest for site survey"
author: "Claudine Gravel-Miguel"
date: "3/21/2022"
output: html_document
---

In this document, I train a random forest model using Spanish archaeological sites to identify where sites might be in Liguria.

As predictors, I use the popularity map created by the AB_LCP, a DEM, the distance from rivers and coast, and the geology at documented caves.

```{r setup, include = F}

library(tidyverse)
library(here) # will help with the paths
library(sp)
library(rgdal)
library(raster)

knitr::opts_chunk$set(warning=F, message=F, echo =F) # Setting up the chunk settings for all chunks

```

```{r setup the projections}

# Setup some projections
crs_lat_long <- 'GEOGCS["GCS_WGS_1984",DATUM["D_WGS_1984",SPHEROID["WGS_1984",6378137,298.257223563]],PRIMEM["Greenwich",0],UNIT["Degree",0.017453292519943295]]'

crs_spain <- 'PROJCS["WGS_1984_UTM_Zone_30N",GEOGCS["GCS_WGS_1984",DATUM["D_WGS_1984",SPHEROID["WGS_1984",6378137.0,298.257223563]],PRIMEM["Greenwich",0.0],UNIT["Degree",0.0174532925199433]],PROJECTION["Transverse_Mercator"],PARAMETER["False_Easting",500000.0],PARAMETER["False_Northing",0.0],PARAMETER["Central_Meridian",-3.0],PARAMETER["Scale_Factor",0.9996],PARAMETER["Latitude_Of_Origin",0.0],UNIT["Meter",1.0]]'

crs_liguria <- 'PROJCS["WGS_1984_UTM_Zone_32N",GEOGCS["GCS_WGS_1984",DATUM["D_WGS_1984",SPHEROID["WGS_1984",6378137.0,298.257223563]],PRIMEM["Greenwich",0.0],UNIT["Degree",0.0174532925199433]],PROJECTION["Transverse_Mercator"],PARAMETER["False_Easting",500000.0],PARAMETER["False_Northing",0.0],PARAMETER["Central_Meridian",9.0],PARAMETER["Scale_Factor",0.9996],PARAMETER["Latitude_Of_Origin",0.0],UNIT["Meter",1.0]]'

```

# Spanish sites during the transition

For the grant proposal, I already cleaned the Vermeersch v27 database to focus on sites with reliable radiocarbon dates, and I calibrated them. So, here I simply imported it and focused on the Spanish sites of the right time period.

```{r importing the Spanish sites}

# As the R project is in the top LiguriaProject folder, with the 'here' package, I can call any file in its subfolders.

european_radiocarbon_db <- read_csv(here("../Application/code/cleaned_Oxcal_cal.csv")) %>%
  filter(date_pm < 300, # focusing on dates with lower error ranges
         between(age.start, -12000, -8000) | between(age.end, -12000, -8000)) %>% 
  distinct(g_coord_long, g_coord_lat, .keep_all = T) %>% 
  distinct(g_sitename, .keep_all = T) # And keep only one instance of each site (need to filter by coordinates and by name to catch most typos)

spanishSites <- european_radiocarbon_db %>%   # the .. allows to go one folder up.
  filter(g_country.y == "Spain",
         g_coord_long < -2, # some sites had completely off coordinates
         g_coord_lat > 41) 

# Import the polygon of the region considered here. This will be used to filter out the points we don't need.
SpainPolygon <- readOGR("../../GISdata/Cantabria_Buffer.shp", verbose = F)
crs(SpainPolygon) <- crs_lat_long
r <- raster(ncol=180, nrow=180)
extent(r) <- extent(SpainPolygon)
SpainPolygonRas <- rasterize(SpainPolygon, r)

spanishSites$Polygon <- extract(x = SpainPolygonRas, 
                                y = SpatialPoints(cbind(spanishSites$g_coord_long,
                                                        spanishSites$g_coord_lat)))

# Remove the sites outside the polygon
spanishSites <- spanishSites %>% 
  filter(!is.na(Polygon))
  
ggplot(spanishSites) +
  geom_point(aes(x=g_coord_long, y=g_coord_lat))

# cleaning up
rm(r, SpainPolygon)

```

# Known Italian sites during the transition

```{r importing the Italian sites}

# As the R project is in the top LiguriaProject folder, with the 'here' package, I can call any file in its subfolders.

italianSites <- european_radiocarbon_db %>% 
  filter(g_country.y == "Italy",
         g_coord_long < 20, # some sites had completely off coordinates
         g_coord_lat > 36) 

ItalyPolygon <- readOGR("../../GISdata/Italian_Polygon.shp", verbose = F)
crs(ItalyPolygon) <- crs_lat_long
r <- raster(ncol=180, nrow=180)
extent(r) <- extent(ItalyPolygon)
ItalyPolygonRas <- rasterize(ItalyPolygon, r)

italianSites$Polygon <- extract(x = ItalyPolygonRas, 
                                y = SpatialPoints(cbind(italianSites$g_coord_long,
                                                        italianSites$g_coord_lat)))

# Remove the sites outside the polygon
italianSites <- italianSites %>% 
  filter(!is.na(Polygon))

ggplot(italianSites) +
  geom_point(aes(x=g_coord_long, y=g_coord_lat))

# cleaning up
rm(r, ItalyPolygon)

```

# Importing caves shapefiles

I obtained the location of caves from the OpenStreetMap project. I am first cleaning the datasets and formatting it in the same way as the archaeological sites datasets. Then, I will join the two so that caves with archaeology have the MesoArch variable as TRUE and the caves who do not (or for which we do not have known presence) will have the variable as FALSE.

To make sure that my join actually links the same caves even if there are typos, I'm simplifying the site names and using those names for the join.

```{r importing shapefiles, results = "hide"}

# As this is a open source dataset, there can be duplicates and typos, so I need to clean the data first
cleanOSM <- function(path){
  
  ds <- data.frame(readOGR(path), verbose = F) # read the shapefile and converts to dataframe
  trial <- ds %>% 
    filter(!is.na(name)) %>% # remove the caves without names
    distinct(name, .keep_all = T) %>% # keep only one of duplicates
    dplyr::select(name=name, desc=descriptio, historic, lat=coords.x2, long=coords.x1)
  
}

spanishCaves <- cleanOSM("../../GISdata/Caves/CantabriaCavesFiltered.shp")
ligurianCaves <- cleanOSM("../../GISdata/Caves/LiguriaCavesFiltered.shp")

```

```{r only caves within polygons}

ligurianCaves$Polygon <- extract(x = ItalyPolygonRas, 
                                y = SpatialPoints(cbind(ligurianCaves$long,
                                                        ligurianCaves$lat)))

# Remove the sites outside the polygon
ligurianCaves <- ligurianCaves %>% 
  filter(!is.na(Polygon))

spanishCaves$Polygon <- extract(x = SpainPolygonRas, 
                                y = SpatialPoints(cbind(spanishCaves$long,
                                                        spanishCaves$lat)))

# Remove the sites outside the polygon
spanishCaves <- spanishCaves %>% 
  filter(!is.na(Polygon))

# no need of the polygons anymore
rm(ItalyPolygonRas, SpainPolygonRas)

```

```{r functions to simplify names}

library(stringi)

wordstoremove <- c("Arma ", "Grotte ", "Abri ", "Grotta ", "Abrigo ", "du ", "de ", "del ", "degli ", "della ", "delle ", "la ", "des ", "à ", "l' ", "Aven ", "Trou ", "Cueva ", "el ", "las ", "los ", "Torca ", "Sima ", "Cuevas ", "en ", "Sima-", "do ", "di ", "dei ", "da ", "dell'", "d'", "a ", "-", " I", " II", " III", " IV", " V", " VI", " VII", " VIII", " IX", " X" , " inf", ", ingresso est", ", ingresso ovest", "ingresso nord", "ingresso sud", "Caverna ", "dello", "Grotticella", "Inferiore", "Superiore", "Pozzetto", "Pozzo", "Riparo", "Sprugola", "Tana ", "Tanna ", "Tanazza ", "Tann-a ", "- ")

cleaningFun <- function(ds){
  
  ds %>% 
    stri_trans_general("Latin-ASCII") %>%  # removes accents
    str_remove_all(fixed("(")) %>% # Remove "("s
    str_remove_all(fixed(")")) %>% # Remove ")"s
    str_remove_all(regex(str_c("\\b",wordstoremove, "\\b", collapse = '|'), ignore_case = T)) %>% 
    str_remove_all("\\.") %>% 
    str_replace_all("[:digit:]", "") %>% # removes any numbers
    str_to_title() %>% # Title case
    str_remove_all(" ") # remove spaces
  
}

addCleanedNames <- function(ds){
  
  ds %>% 
    mutate(nameCleaned = nameNoParen) %>%  # append that clean vector to the dataset
    arrange(nameCleaned) # sort
  
}

```

```{r converting names to simple forms in Spain}

# For caves
nameNoParen <- spanishCaves$name %>%
  cleaningFun()

spanishCaves <- spanishCaves %>% 
  addCleanedNames() %>% 
  distinct(nameCleaned, .keep_all = T) # removing duplicates

# For sites
nameNoParen <- spanishSites$g_sitename %>%
  cleaningFun()

spanishSites <- spanishSites %>% 
  addCleanedNames() %>% 
  distinct(nameCleaned, .keep_all = T) # removing duplicates

write_csv(spanishSites, "../../GISdata/Spanish_Mesolithic_Sites.csv") # I cleaned a few coordinates in QGIS

```

```{r merge caves and sites in Spain}

# Get the Caves without archaeology (even if I delete some wrong names because of duplicates, it doesn't matter much given the number of caves)

spanishSites <- read_csv("../../GISdata/Spanish_Mesolithic_Sites_Cleaned.csv") # reset as I adjusted some coordinates

spanishNoArch <- spanishCaves %>% 
  anti_join(spanishSites, by = "nameCleaned")

# Format so I can merge the two datasets together
spanishNoArch <- spanishNoArch %>% 
  dplyr::select(name, nameCleaned, lat, long) %>% 
  mutate(MesoArch = FALSE)

spanish <- spanishSites %>% 
  dplyr::select(name = g_sitename, nameCleaned, lat = g_coord_lat, long = g_coord_long) %>% 
  mutate(MesoArch = TRUE) %>% 
  rbind(spanishNoArch)

# cleanup 
rm(spanishCaves, spanishSites, spanishNoArch)

```

```{r merge caves and sites in Italy}

# For caves
nameNoParen <- ligurianCaves$name %>%
  cleaningFun()

ligurianCaves <- ligurianCaves %>% 
  addCleanedNames() %>% 
  distinct(nameCleaned, .keep_all = T) # removing duplicates

# For sites
nameNoParen <- italianSites$g_sitename %>%
  cleaningFun()

italianSites <- italianSites %>% 
  addCleanedNames()

write_csv(italianSites, "../../GISdata/Italian_Mesolithic_Sites.csv") # I cleaned a few coordinates in QGIS

italianSites <- read_csv("../../GISdata/Italian_Mesolithic_Sites_Filtered.csv") # reset as I adjusted some coordinates and filtered the dataset by the Liguria shapefile

liguriaNoArch <- ligurianCaves %>% 
  anti_join(italianSites, by = "nameCleaned")

# Format so I can merge the two datasets together
liguriaNoArch <- liguriaNoArch %>% 
  dplyr::select(name, nameCleaned, lat, long) %>% 
  mutate(MesoArch = FALSE)

liguria <- italianSites %>% 
  dplyr::select(name = g_sitename, nameCleaned, lat = g_coord_lat, long = g_coord_long) %>% 
  mutate(MesoArch = TRUE) %>% 
  rbind(liguriaNoArch)

# cleanup 
rm(ligurianCaves, italianSites, liguriaNoArch)

```

# Import the predictor maps

## Import the mobility maps

In this section, I import the rasters created by the AB-LCP model and compile their results by adding their values (sum, ignoring NA).

This map show the normalized summed mobility in Spain.

```{r mobility in Spain}

library(raster)
  
# Create a list of all the file names in the identified path folder
all_files <- list.files(path = "../LCP_Outputs/modelOutputsSpain/", all.files = TRUE, full.names = TRUE, pattern = "\\.asc$")
  
all_maps <- stack(all_files)

sum_mobility_spain <- sum(all_maps, na.rm = T)

# standardize to 0-100
sum_mobility_spain <- (sum_mobility_spain * 100) / maxValue(sum_mobility_spain)

plot(sum_mobility_spain)

rm(all_maps, all_files)

```

This map show the normalized summed mobility in Liguria.

```{r mobility in Liguria}
  
# Create a list of all the file names in the identified path folder
all_files <- list.files(path = "../LCP_Outputs/modelOutputsLiguria/", all.files = TRUE, full.names = TRUE, pattern = "\\.asc$")
  
all_maps <- stack(all_files)

sum_mobility_liguria <- sum(all_maps, na.rm = T)

sum_mobility_liguria <- (sum_mobility_liguria * 100) / maxValue(sum_mobility_liguria)
plot(sum_mobility_liguria)

rm(all_maps, all_files)

```

Those are then reprpjected to LatLong to fit the rest of the data.

```{r reproject mobility and dem maps to Lat Long}

reproject_to_Lat_long <- function(ras){
  
  projectRaster(ras, crs = crs_lat_long)
  
}

crs(sum_mobility_spain) <- crs_spain
sum_mobility_spain <- reproject_to_Lat_long(sum_mobility_spain)
sum_mobility_spain <- round(sum_mobility_spain, 0)

# now for Liguria
crs(sum_mobility_liguria) <- crs_liguria
sum_mobility_liguria <- reproject_to_Lat_long(sum_mobility_liguria)
sum_mobility_liguria <- round(sum_mobility_liguria, 0)

```

## Import the other predictors

Importing the maps of elevation, slope, distance from rivers, distance from coast, viewshed from summits, and geology. Formatting those to make sure that they are in the correct projection (Lat Long).

```{r upload predictors}

# the lat long DEM for both regions is the GMTED clipped file

dem <- raster("../../GISdata/Merged_GMTED_GEBCO_Clipped_Sea_level.tif")
slope <- raster("../../GISdata/Slope.tif")
slope <- round(slope, 0) # rounding out decimals

# dist from rivers (using r.walk in GRASS)
liguria_dist_from_rivers <- raster("../../GISdata/Liguria_rWalk_from_rivers_1000.tif")
liguria_dist_from_rivers <- reproject_to_Lat_long(liguria_dist_from_rivers)
liguria_dist_from_rivers <- round(liguria_dist_from_rivers / 60, 0) # transforms the data in minutes

spain_dist_from_rivers <- raster("../../GISdata/Spain_rWalk_from_rivers_1000.tif")
spain_dist_from_rivers <- reproject_to_Lat_long(spain_dist_from_rivers)
spain_dist_from_rivers <- round(spain_dist_from_rivers / 60, 0) # transforms the data in minutes

# dist from coast (using r.walk in GRASS)
liguria_dist_from_coast <- raster("../../GISdata/Liguria_rWalk_from_coast.tif")
liguria_dist_from_coast <- reproject_to_Lat_long(liguria_dist_from_coast)
liguria_dist_from_coast <- round(liguria_dist_from_coast / 60, 0) # transforms the data in minutes

spain_dist_from_coast <- raster("../../GISdata/Spain_rWalk_from_coast.tif")
spain_dist_from_coast <- reproject_to_Lat_long(spain_dist_from_coast)
spain_dist_from_coast <- round(spain_dist_from_coast / 60, 0) # transforms the data in minutes

# viewshed from summits
spain_viewshed_from_summits <- raster("../../GISdata/Spain_View_from_Summits_norm.tif")
spain_viewshed_from_summits <- reproject_to_Lat_long(spain_viewshed_from_summits)
spain_viewshed_from_summits <- round(spain_viewshed_from_summits, 2)

liguria_viewshed_from_summits <- raster("../../GISdata/Liguria_View_from_Summits_norm.tif")
liguria_viewshed_from_summits <- reproject_to_Lat_long(liguria_viewshed_from_summits)
liguria_viewshed_from_summits <- round(liguria_viewshed_from_summits, 2)

# And some shapefiles
geology <- readOGR(dsn = "../../GISdata/Geological_map_USGS.shp", verbose = F)
geology <- geology['GLG'] # focusing only on the geology type

```

# Extracting raster values at caves' locations

Here, we extract the values from the imported rasters (predictors) using the coordinates of the caves cleaned up earlier.

For mobility and visibility from summit, the etxracting takes the max value within a 1 km radius around the point. For all other values, the value is at the point location.

```{r set two maps to same extent}

library(sp) # for extraction from shapefile

extractRasterValues <- function(ras, ds){
  
  extract(ras, ds)
  
}

liguria$mobility <- extract(sum_mobility_liguria, liguria[,c(4,3)], buffer = 1000, fun = max)
liguria$dem <- extractRasterValues(dem, liguria[,c(4,3)])
liguria$slope <- extractRasterValues(slope, liguria[,c(4,3)])
liguria$dist_rivers <- extractRasterValues(liguria_dist_from_rivers, liguria[,c(4,3)])
liguria$dist_coast <- extractRasterValues(liguria_dist_from_coast, liguria[,c(4,3)])
liguria$visibility <- extract(liguria_viewshed_from_summits, liguria[,c(4,3)], buffer = 1000, fun = max)

pts <- SpatialPoints(liguria[,c(4,3)], proj4string = CRS(proj4string(geology)))
liguria$geology <- over(pts, geology)

spanish$mobility <- extract(sum_mobility_spain, spanish[,c(4,3)], buffer = 1000, fun = max)
spanish$dem <- extractRasterValues(dem, spanish[,c(4,3)])
spanish$slope <- extractRasterValues(slope, spanish[,c(4,3)])
spanish$dist_rivers <- extractRasterValues(spain_dist_from_rivers, spanish[,c(4,3)])
spanish$dist_coast <- extractRasterValues(spain_dist_from_coast, spanish[,c(4,3)])
spanish$visibility <- extract(spain_viewshed_from_summits, spanish[,c(4,3)], buffer = 1000, fun = max)

pts <- SpatialPoints(spanish[,c(4,3)], proj4string = CRS(proj4string(geology)))
spanish$geology <- over(pts, geology)

rm(pts)

```

Here is a quick summary of the data. The first table shows the mean values for Spanish caves, and the second shows the mean for Ligurian caves.

```{r overview of data}

spanish %>% 
  dplyr::select(mobility, dem, slope, distRivers = dist_rivers, distCoast = dist_coast, visibility) %>% 
  summarize(across(everything(), list(min = min, mean = mean, max = max))) %>% 
  gather(key = "key", value = "value") %>%
  separate(key, c("variable", "stat"), sep = "_") %>%
  pivot_wider(names_from = "stat", values_from = "value") 

liguria %>% 
  dplyr::select(mobility, dem, slope, distRivers = dist_rivers, distCoast = dist_coast, visibility) %>% 
  summarize(across(everything(), list(min = min, mean = mean, max = max))) %>% 
  gather(key = "key", value = "value") %>%
  separate(key, c("variable", "stat"), sep = "_") %>%
  pivot_wider(names_from = "stat", values_from = "value") 


```

This is actually telling me that there are values in Liguria (mobility and slope) that will not be encountered during training, which may be a problem as RF does not like that.

# Random Forest

I am training the random forest algorithm on the Spanish dataset to see if we can predict the presence of Mesolithic archaeology at sites

```{r random forest algorithm}

library(randomForest)
library(pROC)

runningRFmodel <- function(ds, trainSampSize, nTrees){
  
  # Dividing into training and test sets
  # Determine the number of rows for training
  n <- nrow(ds)
  
  # Create a random sample of row IDs
  sample_rows <- sample(n, n*trainSampSize)
  
  # Create the training dataset
  ds_train <- ds[sample_rows,]
  
  # Create the test dataset with the rows that are NOT in the training dataset
  ds_test <<- ds[-sample_rows,]
  
  # Build a random forest model
  ds_model <- randomForest(MesoArch~., data = ds_train, ntree = nTrees, type = "classification", na.action = na.omit)
  
  # test the model on the test dataset
  ds_test$pred <<- predict(ds_model, ds_test, type = "response")
  
  # Compute the accuracy of the random forest
  print(paste0("Mean accuracy on test data: ", mean(ds_test$pred == ds_test$MesoArch)))
  
  # Compute the ROC curve and AUC
  rf.roc <- roc(ds_train$MesoArch, ds_model$votes[,2])
  plot(rf.roc)
  print(auc(rf.roc))
 
  return(ds_model)
}

```

Using seed 12, with 5000 trees and 50/50 training vs test dataset.

```{r running the RF}

set.seed(12) # for reproducibility
# I actually tried a few seeds and 2, 8 give OK results for Spain, but most importantly, gets all three known sites in Liguria correctly. But those predict 100+ sites in Liguria. 12 is the best seed to use (predicts well AC, AV, and AS, but only identifies 63 more possible sites).

spanish_full_for_RF <- spanish %>% 
  dplyr::select(dem, mobility, slope, dist_rivers, dist_coast, visibility, geology, MesoArch) %>%
  mutate(geology = unlist(geology)) %>% 
  drop_na() %>% 
  mutate(MesoArch = factor(MesoArch))

minCat <- spanish_full_for_RF %>% 
  count(MesoArch)

minCat <- min(minCat$n)

# Resample with replacement
spanish_for_RF <- spanish_full_for_RF %>%
  group_by(MesoArch) %>%
  sample_n((minCat*4), replace=T) %>% 
  ungroup()

ds_model <- runningRFmodel(spanish_for_RF, 0.50, 5000)

data.frame(ds_model$importance) %>% 
  arrange(desc(MeanDecreaseGini))

ds_model$confusion

```

When testing the model on the full dataset, it is pretty good. This is the percentage of caves that were correctly predicted:

```{r test on full Spain dataset}

spanish_full <- spanish %>% 
  mutate(geology = unlist(geology)) %>%
  drop_na()

spanish_full$pred <- predict(ds_model, spanish_full_for_RF)

sum(spanish_full$MesoArch == spanish_full$pred)/nrow(spanish_full)

spanish_full <- spanish_full %>% 
  mutate(geology = unlist(geology)) %>% 
  mutate(prediction_type = ifelse(MesoArch == TRUE,
                                  ifelse(pred == TRUE, "TruePositive", "FalseNegative"),
                                  ifelse(pred == FALSE, "TrueNegative","FalsePositive")))

write_csv(spanish_full, "../../GISdata/Spain_predicted.csv")

```

Applying the model to the Ligurian dataset, it predicts accurately the presence of Meso archaeology in the 3 sites with known archaeology from that period. This is satisfying for now.

```{r apply model on Ligurian caves}

liguria_for_RF <- liguria %>% 
  dplyr::select(dem, mobility, slope, dist_rivers, dist_coast, geology, visibility) %>% 
  mutate(geology = unlist(geology)) %>% 
  filter(!is.na(dem))

# test the model on the test dataset
liguria$pred <- predict(ds_model, liguria_for_RF, type = "response")

liguria %>% 
  as_tibble() %>% 
  dplyr::select(c(name, MesoArch, pred))

# How many sites with Meso arch predicted?
sum(liguria$pred == TRUE)

```

According to this model, these are the sites that we should look into for our surveys:

```{r survey sites}

liguria %>% 
  filter(pred == TRUE) %>% 
  dplyr::select(name) %>% 
  arrange(name)

liguria %>% 
  ggplot() +
  geom_point(aes(x=long, y=lat, color = pred))

liguria <- liguria %>% 
  mutate(geology = unlist(geology))

write_csv(liguria, "../../GISdata/Liguria_predicted.csv")

```

That is a lot of sites, but fortunately, most of them are close to one another. We should expect a fair number of false positives, however.

# The end
